Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-11-30 responseToReviews.md

> This is a good point and we now include the Boyd reference in the section on fingerprinting.

## Reviewer: 2

# Comments to the Author
This manuscript is intended to be a review of the field of steady-state visual evoked potentials (ssVEPs) over the past 70 years.  The authors state in the introductory paragraph “Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.” The material included and the references cited, however, are quite limited and omit seminal work in this area.

> We thank the reviewer for their helpful comments. As we noted above, our intention here was to provide an introduction to the SSVEP technique with appropriate simple code samples that would allow relative newcomers to the field to implement the the technique. We are, however, keen to avoid misrepresenting the history of SSVEP research and are happy to include appropriate references.

@@

> Thank you for this explanation. We have now reworded the paper to make this clearer. We did take care to define what we mean by 'on-off' flicker in the preceding sentence but we now clarify the alternative 'appearence/disappearence' terminology. The point we were making was that odd harmonic responses (1F) are a feature of this this type of modulation because the responses to the appearance and disappearance of a pattern (or mean field) need not be completely symmetrical (See Zemon, et al 1988 for an explicit demonstration).

This type of stimulation does not separate contributions from “independent populations of on- and off-cells.”   There are other ssVEP techniques developed that do tap ON and OFF pathways selectively, which are not cited here. Contrast-reversal stimuli which involve symmetrical modulation of light and dark elements cancel odd harmonic frequency components in the response, however, both kinds of stimulus techniques mentioned here elicit only nonlinear responses.

> We have clearly not been sufficiently clear here. We do not think we used the term 'separation' and we agree that a) there are infrequently-used paradigms that can target ON and OFF pathways more effectively (for example, 'sawtooth' temporal modulations that provide asymmetrical onset or offset transients) and b) responses from spatial features of differing polarity will tend to be combined at the scalp as long as the spatial frequency is high enough. We have expanded this section to attempt to make the general point clear.

@@

3. The statement “SSVEPs have proven to be an excellent measure of early chromatic processing as well” again fails to cite seminal work in this area.

> We have included three earlier papers by Regan on this topic.

4. The statement “This normalization, achieved through a computation called ‘contrast gain control’” again neglects to cite the original work that defined “contrast gain control” and the authors ignored a nonlinear biophysical model proposed to explain this phenomenon in ssVEP contrast response functions (which was also applied in developmental work on infants and children).

> Contrast gain control is related to 'automatic gain control' - a electrical engineering concept that arose in the early 20th century. The concept of modulating the input (rather than the output) of the system became established in the 1930s and 40s (for example in WW2 radar systems) and the specific idea of computing that modulation as a function of the local average signal power seems to have arisen around this time and was refined over the 50s and 60s (for example, in 'sidechain compression'). In the domain of vision, an early example of 'contrast gain control' (scaling the input - in this case luminance) is found in Sperling and Sondhi 1968 but the first formalization of the concept in the domain of spatial vision that included the critical concepts of an early, divisive gain control signal computed from some additional estimate of local contrast seems to appear in Heeger 1992 (to explain animal data) and Foley 1994 where it is applied to human psychophysical data.

1. The description of Figure 3 and the nonlinear operations performed are unclear. They state in the text that Equation 2 (“hyperbolic ratio function”) is used as a nonlinear transducer, but the figure legend states that a squaring operator was used for frequency doubling. A squaring operation would only produce a sinusoid of double the frequency of the input. The waveform shown in the left panel appears to be a full-wave rectified signal, which would contain theoretically an infinite number of higher harmonics. The text also states that “This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity.” When was full-wave rectification performed?

### 2024-12-01 responseToReviews.md

3. The statement “SSVEPs have proven to be an excellent measure of early chromatic processing as well” again fails to cite seminal work in this area.

> We have included three earlier papers by Regan on this topic.

4. The statement “This normalization, achieved through a computation called ‘contrast gain control’” again neglects to cite the original work that defined “contrast gain control” and the authors ignored a nonlinear biophysical model proposed to explain this phenomenon in ssVEP contrast response functions (which was also applied in developmental work on infants and children).

> Contrast gain control is related to 'automatic gain control' - a electrical engineering concept that arose in the early 20th century. The concept of modulating the input (rather than the output) of the system became established in the 1930s and 40s (for example in WW2 radar systems). The specific idea of computing that modulation as a function of the local average signal power seems to have arisen around this time and was refined over the 50s and 60s (for example, in 'sidechain compression'). In the domain of vision, an early example of 'contrast gain control' (scaling the input - in this case luminance) is found in Sperling and Sondhi 1968.

The first formalization of the concept in the domain of spatial vision that included the critical concepts of an early, divisive gain control signal computed from some additional estimate of local contrast seems to appear in   a seminal paper by Shapley and Victor, 1978. Application to contrast adaptation was demonstrated by Ohzawa, Sclar, Freeman, 1982, 1985] with a demonstration of instantaneous gain control in the human SSVEP by Bobak et al in 1988. A synthesis that included both adaptation and masking was proposed by Heeger 1992 while Foley 1994 show an application to human psychophysical data. We have now cited these more foundational papers. We would be happy to include the references to the non-linear biophysical model used in adult and developmental work if the reviewer can point us to the appropriate literature.

@@

1. The description of Figure 3 and the nonlinear operations performed are unclear. They state in the text that Equation 2 (“hyperbolic ratio function”) is used as a nonlinear transducer, but the figure legend states that a squaring operator was used for frequency doubling. A squaring operation would only produce a sinusoid of double the frequency of the input. The waveform shown in the left panel appears to be a full-wave rectified signal, which would contain theoretically an infinite number of higher harmonics. The text also states that “This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity.” When was full-wave rectification performed?

> The transducer function used to produce Figure 3 is available in the paper and reproduced here:

```python
def simpleTransducerFunction(inputContrast, c50, RMax):
 # Here we model a simple hyperbolic ratio function. Also called a "Naka-Rushton" function.
 # RMax defines the reponse gain of the transducer (how big the biggest output is). 
 # c50 defines the conrast sensitivity. Smaller values are more sensitive.
 # The exponent can also vary - here we fix it to 2 but it can be varied.
  expnt=2

  # Contrast cannot be negative. To model the responses of a set of on and off 
  # contrast detectors we perform full-wave rectification of the signal
  inputContrast=np.abs(inputContrast)
  output=RMax*(inputContrast**expnt)/(inputContrast**expnt + c50**expnt)
  return output

```

The input contrast is full-wave rectified (via the 'abs' operator). Here we choose an exponent of '2' for the hyperbolic ratio function so it is *also* squared both in the numerator and the denominator. But this parameter is often fitted rather than fixed and the reader is encouraged to experiment with the exponent to examine its effect on the resulting amplitude spectrum. While there are, technically, infinite harmonics of this input, they decrease with frequency and we plot only three for reasons of space.

### 2024-12-01 review.qmd
Adding figure for full wave rectification

def simpleTransducerFunction(inputContrast, c50, RMax):
 # Here we model a simple hyperbolic ratio function. Also called a "Naka-Rushton" function.
 # RMax defines the reponse gain of the transducer (how big the biggest output is). 
 # c50 defines the conrast sensitivity. Smaller values are more sensitive.
 # The exponent can also vary - here we fix it to 2
  expnt=2

  # Contrast cannot be negative. To model the responses of a set of on and off 
  # contrast detectors we perform full-wave rectification of the signal
  inputContrast=np.abs(inputContrast)
  output=RMax*(inputContrast**expnt)/(inputContrast**expnt + c50**expnt)
  return output

@@

initialSignal = np.zeros(len(contrastRange)) 
transducedSignal = np.zeros(len(contrastRange))
for contrastIndex,thisContrast in enumerate(contrastRange):
  inputModulation = np.sin(2 * np.pi * inputFrequency_Hz * support_S) * thisContrast

@@

```{python transducerout}

@@

@@

axs[2].grid()
plt.tight_layout()
plt.savefig('Figures/3_transducer.pdf')

@@

axs[2].grid()
plt.tight_layout()
plt.savefig('Figures/4_twoinputs.pdf')

@@

axs.set_title(f'Simulated sweep VEP at {nLevels} different contrasts', fontsize=10)
axs.semilogx(contrastRange, transducedSignal, 'o')
axs.semilogx(contrastRange[:4], reg.predict(X), 'r-', label='Linear Fit')
axs.semilogx(np.logspace(0.1,1.1,nLevels)/100, reg.predict(X2), 'r--')
axs.semilogx([x_intercept], [0], 'ro', label='Threshold')
axs.set_xlabel('Contrast')
axs.set_ylabel('Amplitude (a.u.)')
axs.set_xlim(0.01, 1)
axs.set_ylim(0, 10)
axs.legend()
axs.grid()
plt.tight_layout()
plt.savefig('Figures/5_sweepplot.pdf')

@@

# To save the figure
plt.savefig('Figures/6_clinicalplot.pdf')

@@

Next we can consider the impact of a nonlinearity in processing on the responses. One of the simplest nonlinearities is the function that describes a cell's response to different levels of contrast. In @fig-transducer this is modeled by a hyperbolic ratio function resulting in a saturating non-linearity:

@@

![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input wavefore (left panel) is frequency doubled by the squaring operator, and now features energy at the second harmonic (10Hz) and its multiple (middle panel). The contrast response function at the second harmonic is now nonlinear, and saturates at high contrasts (right panel).](){#fig-transducer}
:::
::: {.content-visible unless-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input wavefore (left panel) is frequency doubled by the squaring operator, and now features energy at the second harmonic (10Hz) and its multiple (middle panel). The contrast response function at the second harmonic is now nonlinear, and saturates at high contrasts (right panel).](Figures/3_transducer.pdf){#fig-transducer}
:::

>>>>

def simpleTransducerFunction(inputContrast, c50, RMax):
 # Here we model a simple hyperbolic ratio function. Also called a "Naka-Rushton" function.
 # RMax defines the reponse gain of the transducer (how big the biggest output is). 
 # c50 defines the conrast sensitivity. Smaller values are more sensitive.
 # The exponent can also vary - here we fix it to 2. Varying it produces different amplitude spectra.

  expnt=2

  output=RMax*(inputContrast**expnt)/(inputContrast**expnt + c50**expnt) # Here we apply the hyperbolic ratio function

  return output

@@

initialSignal = np.zeros(len(contrastRange)) 
transducedSignal = np.zeros(len(contrastRange))

for contrastIndex,thisContrast in enumerate(contrastRange):
  inputModulation = np.sin(2 * np.pi * inputFrequency_Hz * support_S) * thisContrast

@@

```
```{python fullwaverectification}

@@

  # We are about to model response to contrast. Different cell populations detect positive and negative-going contrast and we can model their combined response by performing full-wave rectification as follows: 
  fullWaveRectifiedSignal=np.abs(inputModulation)

  # This doubles the frequencies of the resulting FT and also introduces harmonics at higher frequencies but it does not alter the general shape of the contrast vs response function.

   # Now compute the FT on the signal after it has passed through full wave rectification
  FT_transducedModulation=np.fft.fft(fullWaveRectifiedSignal)/ len(fullWaveRectifiedSignal)
  transducedAmplitude = np.abs(FT_transducedModulation[inputFrequency_Hz*2])
  transducedSignal[contrastIndex]=transducedAmplitude[0]

@@

axs[2].grid()
plt.tight_layout()
plt.savefig('Figures/3_fullwaverectification.pdf')

```
```{python transducerout}

initialSignal = np.zeros(len(contrastRange)) 
transducedSignal = np.zeros(len(contrastRange))

for contrastIndex,thisContrast in enumerate(contrastRange):
  inputModulation = np.sin(2 * np.pi * inputFrequency_Hz * support_S) * thisContrast

  # We are about to model response to contrast. Different cell populations detect positive and negative-going contrast and we can model their combined response by performing full-wave rectification as follows: 
  fullWaveRectifiedSignal=np.abs(inputModulation)

  # This doubles the frequencies of the resulting FT and also introduces harmonics at higher frequencies but it does not alter the general shape of the contrast vs response function.

   # Now compute the FT on the signal after it has passed through a saturating non-linearity. Do this 
   # using the transducer function above.
  transducedModulation=simpleTransducerFunction(fullWaveRectifiedSignal, c50,RMax)
  FT_transducedModulation=np.fft.fft(transducedModulation)/ len(transducedModulation)
  transducedAmplitude = np.abs(FT_transducedModulation[inputFrequency_Hz*2])
  transducedSignal[contrastIndex]=transducedAmplitude[0]

axs[2].grid()
plt.tight_layout()
plt.savefig('Figures/4_transducer.pdf')

@@

axs[2].grid()
plt.tight_layout()
plt.savefig('Figures/5_twoinputs.pdf')

@@

axs.set_title(f'Simulated sweep VEP at {nLevels} different contrasts', fontsize=10)
axs.semilogx(contrastRange, transducedSignal, 'o')
axs.semilogx(contrastRange[:4], reg.predict(X), 'r-', label='Linear Fit')
axs.semilogx(np.logspace(0.1,1.1,nLevels)/100, reg.predict(X2), 'r--')
axs.semilogx([x_intercept], [0], 'ro', label='Threshold')
axs.set_xlabel('Contrast')
axs.set_ylabel('Amplitude (a.u.)')
axs.set_xlim(0.01, 1)
axs.set_ylim(0, 10)
axs.legend()
axs.grid()
plt.tight_layout()
plt.savefig('Figures/6_sweepplot.pdf')

@@

# To save the figure
plt.savefig('Figures/7_clinicalplot.pdf')

@@

Next we can consider the impact of two types of nonlinearity in processing on the responses.

First, we consider the fact that neurons in the early visual system code either positive or negative contrast. The responses of each different type of neuron is an increase in firing rate to the increasing levels of the prefered contrast polatiry and these neurons typically exhibit zero (or baseline) response to the non-prefered polarity. We can model the combined responses of these two cell populations to a time varying contrast signal by full-wave-rectification. The effect of this computation is, effectively, to double the input freuquency in the population measurement and to introduce additional higher (even) harmonics due to the discontinuity at the contrast reversals. The effect of full-wave rectification on the SSVEP signal is shown in @fig-fullwaverectification

![Illustration of a sinusoidal input signal (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right). These calculations assume an entirely linear system.](){#fig-fullwaverectification}
:::
::: {.content-visible unless-format="docx"}
![Illustration of a full-wave rectified sinusoid (left), its Fourier spectrum (middle), and how the amplitude of the first harmonic component increases with contrast (right).](Figures/3_fullwaverectification.pdf){#fig-fullwaverectification}
:::

One of the simplest nonlinearities is the function that describes a cell's response to different levels of contrast. In @fig-transducer this is modeled by a hyperbolic ratio function resulting in a saturating non-linearity:

@@

![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input wavefore (left panel) is frequency doubled by the squaring operator, and now features energy at the second harmonic (10Hz) and its multiple (middle panel). The contrast response function at the second harmonic is now nonlinear, and saturates at high contrasts (right panel).](){#fig-transducer}
:::
::: {.content-visible unless-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input wavefore (left panel) is frequency doubled by the squaring operator, and now features energy at the second harmonic (10Hz) and its multiple (middle panel). The contrast response function at the second harmonic is now nonlinear, and saturates at high contrasts (right panel).](Figures/4_transducer.pdf_transducer.pdf){#fig-transducer}
:::

### 2024-12-01 review.tex
Adding figure for full wave rectification

Next we can consider the impact of a nonlinearity in processing on the
responses. One of the simplest nonlinearities is the function that
describes a cell's response to different levels of contrast. In
Figure~\ref{fig-transducer} this is modeled by a hyperbolic ratio
function resulting in a saturating non-linearity:

@@

\includegraphics{Figures/3_transducer.pdf}

>>>>

Next we can consider the impact of two types of nonlinearity in
processing on the responses.

First, we consider the fact that neurons in the early visual system code
either positive or negative contrast. The responses of each different
type of neuron is an increase in firing rate to the increasing levels of
the prefered contrast polatiry and these neurons typically exhibit zero
(or baseline) response to the non-prefered polarity. We can model the
combined responses of these two cell populations to a time varying
contrast signal by full-wave-rectification. The effect of this
computation is, effectively, to double the input freuquency in the
population measurement and to introduce additional higher (even)
harmonics due to the discontinuity at the contrast reversals. The effect
of full-wave rectification on the SSVEP signal is shown in
Figure~\ref{fig-fullwaverectification}

\includegraphics{Figures/3_fullwaverectification.pdf}

\caption{\label{fig-fullwaverectification}Illustration of a full-wave
rectified sinusoid (left), its Fourier spectrum (middle), and how the
amplitude of the first harmonic component increases with contrast
(right).}

One of the simplest nonlinearities is the function that describes a
cell's response to different levels of contrast. In
Figure~\ref{fig-transducer} this is modeled by a hyperbolic ratio
function resulting in a saturating non-linearity:

@@

\includegraphics{Figures/4_transducer.pdf_transducer.pdf}
